Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2025-02-07 report/emse24-revised.tex
chore(report): emse24-revised updates

    To gain a deeper understanding of Python assertions written in Jupter notebooks, we conduct individual case-studies with 82 assert statements. We analyse the assertion along with the surrounding code to understand its purpose. Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

@@

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}. The libraries are selected based on the import analysis conducted by \citet{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub. We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

\subsection{Case Studies}

For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last cell statements. To surface interesting candidates for the case studies, we apply text processing techniques as described below.

@@

Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks. The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\citep{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

Our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle. Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.

We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\citep{heer2010tour}. By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.

@@

The findings of this study align with prior research on real-world defects in ML systems. \citet{humbatova2020taxonomy} propose a taxonomy of faults in DL systems, while \citet{morovati2023bugs} curate \emph{Defects4ML}, a dataset consisting of ML bugs. Both studies identify defects arising from the use of incorrect data types and shapes, deprecated APIs, and missing GPU resources. Our study identifies several explicit assertions that can catch such data errors. Additionally, we find several implicit checks performed using print and last cell statements to verify the presence of compute resources and libraries, monitor model training, and check the architecture of neural networks.

@@

Additionally, future work can also focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.

@@

Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research. We believe that the primary challenge here will be to develop an appropriate definition for notebook complexity.

To the best of our knowledge, no formal definition or metrics to quantify notebook complexity exist to date. In their research, \citet{grotov2022large-scale} investigated the complexity of Python code written inside Jupyter notebooks using a subset of 15 metrics derived from prior literature. These metrics however were primarily developed for code structured in an object-oriented programming paradigm. Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices \cite{pimentel2019large-scale}.

@@

We enforced a time resource limit of 200 hours for conducting the case studies in RQ2 and RQ3, potentially constraining the depth and breadth of our investigation. The stratified random sampling used for selecting the case study candidates mitigates this threat by ensuring that a representative cross-section of the population is sampled within the time constraint. While this sampling method enhances the generalizability of our results, the time limitation should still be considered when interpreting and applying the study's findings.

@@

\paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks publicly available on GitHub and Kaggle. Consequently, the findings may not be generalizable to Python notebooks that do not concentrate on ML. Additionally, the results may not be applicable to notebooks authored in other statistical programming languages (such as R or Julia), or to proprietary notebooks hosted on different platforms.

@@

Despite these advancements, there is a notable gap in practical, day-to-day testing practices within ML development environments. Our study addresses this gap by analysing the feedback mechanisms in Jupyter notebooks, highlighting how assertions and print statements are used to validate data and model performance, and proposing ways to integrate these practices more effectively into the ML development lifecycle.

>>>>

    To gain a deeper understanding of Python assertions written in Jupyter notebooks, we conduct individual case-studies with 82 assert statements. We analyse the assertion along with the surrounding code to understand its purpose. Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

@@

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.  Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}.  The libraries were chosen based on the import analysis conducted by \citet{psallidas2019data}, which analyzed six million Jupyter notebooks collected from GitHub. It is important to note that we excluded libraries such as Pandas, NumPy, and SciPy, since can be utilized for a wide range of tasks beyond machine learning and do not specifically facilitate model training.

We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

\subsection{Exploratory Case Studies}

For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements. To surface interesting candidates for the case studies, we apply text processing techniques as described below.

@@

Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks. The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies. Moreover, our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle. Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\citep{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

% TODO: this paragraph needs work...
We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\citep{heer2010tour}. By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation. The dataset contributed by this study provides a valuable resource to automatically generate documentaion from implicit and explicit feedback mechanisms written inside Jupyter notebooks~\cite{yang2021subtle}.

% Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.

@@

The findings of this study align with prior research on real-world defects in ML systems. \citet{humbatova2020taxonomy} propose a taxonomy of faults in DL systems, while \citet{morovati2023bugs} curate \emph{Defects4ML}, a dataset consisting of ML bugs. Both studies identify defects arising from the use of incorrect data types and shapes, deprecated APIs, and missing GPU resources. Our study identifies several explicit assertions that can catch such data errors. Additionally, we find several implicit checks performed using print and last cell statements to verify the presence of compute resources and libraries, monitor model training, and check the architecture of neural networks. Future work should focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.

@@

@@

% TODO: why is it important to analyze FMs wrt complexity of notebooks?
% TBH I do not agree with the reviewer's comment about this...
Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research. We believe that the primary challenge here will be to develop an appropriate definition for notebook complexity.

To the best of our knowledge, no formal definition or metrics to quantify notebook complexity exist to date. In their research, \citet{grotov2022large-scale} investigated the complexity of Python code written inside Jupyter notebooks using a subset of 15 metrics derived from prior literature. These metrics however were primarily developed for code structured in an object-oriented programming paradigm. Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices~\cite{pimentel2019large-scale}.

@@

We enforced a time resource limit of 200 hours for conducting the case studies in RQ2 and RQ3, potentially constraining the depth and breadth of our investigation. Although alternative stopping rules such as theoretical saturation are common, they are inherently subjective and can vary based on the judgement of a researcher. In contrast, a time-bound method ensures transparency and reflects the practical resource constraints in real-world research. Additionally, the stratified random sampling used for selecting the case study candidates mitigates this threat by ensuring that a representative cross-section of the population is sampled within the time constraint. While this sampling method enhances the generalizability of our results, the time limitation should still be considered when interpreting and applying the study's findings.

@@

\paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks that are publicly available on platforms such as GitHub and Kaggle. As a result, the findings may not be generalizable to Python notebooks that do not center around ML applications. Furthermore, the applicability of the results is limited when it comes to notebooks authored in other statistical programming languages, such as R or Julia, or to proprietary notebooks hosted on different platforms.

The analysis in this study is conducted only using Jupyter notebooks. However, the underlying workflow of users across all notebook environments is identical---that is, users execute chunks of code using a REPL mechanism. As such, the findings of this study may be generalizable to other notebook technologies.

@@

Despite these advancements, there is a notable gap in practical, day-to-day testing practices within ML development environments. To the best of our knowledge, \citet{openja_empirical_2024} is the only study that empirically investigates the test code utilized within 11 ML projects spanning four different application domains. Our study builds upon the findings of \citet{openja_empirical_2024} by further analyzing the feedback mechanisms within Jupyter notebooks. We highlight how assertions and print statements are employed to validate data and model performance, and propose strategies for integrating these practices more effectively into the ML development lifecycle.
